\documentclass[12pt]{article}
\usepackage{bbold}
\usepackage{semantic}
\usepackage{mathtools}
\usepackage{syntax}
\usepackage{xstring}
\usepackage{trfrac}
\usepackage{listings}

\renewcommand{\syntleft}{\normalfont\itshape}
\renewcommand{\syntright}{}

\newcommand{\cmp}[2]{
  \mathcal{#1}|[ \mathtt{ \StrSubstitute{#2}{ }{\ } } |]
}

\newcommand{\syn}[1]{
  \synt{\textbf{#1}}
}

\newcommand{\ul}[1]{
  \underline{#1}
}

\begin{document}
\begin{enumerate}
  \item Questions
    \begin{enumerate}
      \item How can you ``remember'' computation or delay it and then ``cleanup''
      \item Does applying them one at a time change anything?
      \item Is there work on program transformation that applies?
      \item Do monadic computations generally provide checkpoints? Monadic ``loops''?
      \item Instruction reordering an issue for updates?
      \item Is Haskell's \texttt{runState} an \texttt{update} equivalent?
      \item Where does the diff begin end?
      \item Could side effects force the computation?
      \item How can you safely do partial updates?
      \item What can be done without types?
    \end{enumerate}

  \item Notes
    \begin{enumerate}
      \item alpha equivalence is a no-op
      \item Combining ``in-flight'' updates with checkpoint has value when the developer misplaces/forgets the \texttt{update} or the execution takes to long to reach it.
      \item Replacing without the update/timeout requirement removes cognitive overhead.
      \item Prove: handles modules (through let?)
      \item Prove: same update never applied twice
      \item Prove: When updates are added to $\Delta$ they always take place before the term is reduced
      \item Prove: updates only apply to original terms
      \item Should preserve the language semantics
      \item A change requires type checking in old and new environment
    \end{enumerate}

  \item TODO
    \begin{enumerate}
      \item Beta equivalence (undecidable)
      \item $\Delta$ ``cleanup'' for ``unused'' updates
      \item Type checking
      \item Reduced/Partially Applied term updates
      \item Working on impure functions
    \end{enumerate}

\newpage

  \item Pre-flight Updates

    Simple

    \begin{equation*}
      (\lambda z.\lambda x.\lambda y.x)^0\ (\lambda x.x)^{10}
    \end{equation*}

    \begin{equation*}
      \Delta = \{(\lambda x.\lambda y.x,\
      \ul{\lambda n}.\lambda x.\lambda y.x\ \ul{n})^{3} \}
    \end{equation*}

    \begin{equation*}
      (\lambda z.\lambda x.\lambda y.x)^0\ (\lambda x.x)^{10} \rightarrow
      (\lambda x.\lambda y.x)^3 \rightarrow
      (\ul{\lambda n}.\lambda x.\lambda y.x\ \ul{n})^3
    \end{equation*}


    \begin{equation*}
      \trfrac[E-UpdateLambda]{
        (\lambda x.t_1, t)^x \in \Delta \qquad \Delta' = \Delta \setminus (\lambda x.t_1, t)^x
      }{
        (\lambda x.t_1)^x\ t_2 ->
        t\ t_2\
      }
    \end{equation*}

    \begin{equation*}
      \trfrac[E-UpdateBody]{
        (t_1, t)^{x} \in \Delta \qquad \Delta' = \Delta \setminus (t_1, t)^{x}
      }{
        (\lambda x.t_1)^{x}\ t_2 ->
        (\lambda x.t)^{x}\ t_2
      }
    \end{equation*}

    \begin{equation*}
      \trfrac[E-AppBeta]{
        (\lambda x.t_1,\_), (t_1, \_) \notin \Delta
      }{
        (\lambda x.t_1)^{x}\ t_2 ->
        [x \mapsto t_2]\,t_1^{\;x+3 }
      }
    \end{equation*}

    \begin{equation*}
      \trfrac[E-Update]{
        \Delta -> \Delta \cup \Delta'
      }{
        t -> t
      }
    \end{equation*}

  \item Explicit Updates

    \begin{equation*}
      \trfrac[E-Update]{
        \Delta -> \Delta \cup \Delta'
      }{
        \mathtt{update}\ t -> t
      }
    \end{equation*}

  \item Updates for Reduced Terms

    \begin{equation*}
      \trfrac[E-AppBetaPart]{
        dom(\langle \ldots \rangle)(\lambda x.\lambda y.t_1,\_) \notin \Delta
      }{
        \langle \ldots \rangle(\lambda x.\lambda y.t_1)^{x}\ t_2 ->
        \langle \ldots, x \mapsto t_2 \rangle(\lambda y.t_1)^{\;x+3 }
      }
    \end{equation*}

    \begin{equation*}
      \trfrac[E-AppBetaSubApp]{
        (\lambda x.t_1\ t_2,\_) \notin \Delta
      }{
         \langle \ldots \rangle (\lambda x.t_1\ t_2)^{x}\ t ->
         [ \ldots, x \mapsto t ] \,(t_1\ t_2)^{\;x+3 }
      }
    \end{equation*}

    \begin{equation*}
      \trfrac[E-AppBeta]{
        (\lambda x.var,\_) \notin \Delta
      }{
         \langle \ldots \rangle (\lambda x.var)^{x}\ t ->
         [ \ldots, x \mapsto t] var^{\;x+3 }
      }
    \end{equation*}

  \item Pure update example
    \lstinputlisting[language=C]{code/rewind.c}

\end{enumerate}

\end{document}
